ProteInfer: deep networks for predicting functional properties of proteins

### 2021-09-14 latex/draft3.tex
Stage more files

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for predicting functional properties of proteins}
\shorttitle{ProteInfer}

\author[1,\Letter]{Authors}
\affil[1]{Google}

\maketitle

\begin{abstract}
Predicting the function(s) of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. In our work, we instead employ deep convolutional neural networks to directly predict a variety of protein functions - EC numbers and GO terms - directly from an unaligned amino acid sequence. This performs comparably to alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces. In response, we have provided an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract
\begin{keywords}
neural networks | protein | prediction | classification | embedding
\end{keywords}
\begin{corrauthor}
\todo{x} \at google.com
\end{corrauthor}
\section*{Introduction}
Every day, more than a hundred thousand protein sequences are added to global sequence databases \cite{tremblstats}. However, these entries are of limited use to practitioners unless they are accompanied by functional annotations. While curators diligently extract annotations from the literature, assessing more than 60,000 papers each year \cite{uniprotlatest}, the time-consuming nature of this task means that only 0.03\% of publicly available protein sequences are manually annotated.
%%%FIGURETREMBL%%%
The community has a long history of using computational tools to infer protein function directly from amino acid sequence. Starting in the 1980s, methods such as BLAST \cite{blast} relied on pairwise sequence comparisons, where a query protein is assumed to have the same function as highly similar sequences that have already been annotated. Signature-based approaches were later introduced, with the PROSITE database \cite{bairoch1991prosite} cataloguing short amino acid ``motifs'' found in proteins that share a particular function. Subsequently, a crucial refinement of signature-based approaches was the development of profile hidden Markov models (HMMs) \cite{krogh1994hidden, eddyhmmreview}. These models collapse an alignment of related protein sequences into a model that provides likelihood scores for new sequences, that describe how well they fit the aligned set. Critically, profile HMMs allow for longer signatures and fuzzier matching and are currently used to update popular databases such as Interpro and Pfam \cite{pfamlatest, blum2021interpro}. Subsequent refinements have made these techniques more sensitive and computationally efficient \cite{altschul1997gapped, fasterhmmer, pfamlatest, soding2005protein, steinegger2019hh}, while their availability as web tools allows practitioners to easily incorporate them into workflows \cite{johnson2008ncbi, soding2005hhpred, potter2018hmmer, bernhofer2021predictprotein}.
\begin{figure}[t]
  \includegraphics[width=\columnwidth]{Figures/new_the_task.pdf}%
  \caption{\textbf{Predicting functional properties from amino acid sequence has traditionally been achieved by comparing to many sequences (BLAST) or many models (InterProScan), deep learning models provide an alternative approach.}  A single neural network is trained to predict a large number of functional properties from an amino acid sequence.\label{fig:task}}
These computational modelling approaches have had great impact; however, one third of bacterial proteins still cannot be annotated (even computationally) with a function \cite{phenotypes}. It is therefore worthwhile to examine how new approaches might complement existing techniques. First, current approaches conduct entirely separate comparisons for each comparator sequence or model, and thus may not fully exploit the features shared across different functional classes. An ideal classification system, for example, might have a modular ATP-binding region detector used in detection of both kinases and ABC transporters \citep{pmid12468712}. Separately modelling these targets in each family, like standard HMM approaches, increases the computational cost and may also be less accurate. In addition, the process of creating many of these signatures is not fully automated and requires considerable curatorial efforts \citep{pfamprofilebuilding, pfamrepeats}, which at present are spread across an array of disparate but overlapping signature databases \citep{blum2021interpro}.

@@

Of particular relevance to the present work is Bileschi et al. (2019) \citep{bileschi}, where it is shown that models with residual layers \citep{resnet} of dilated convolutions \citep{dilatedconvolutions} can categorise protein domains more accurately than profile HMMs or BLAST. Dohan (2021) \cite{dohan2021kdd} provides additional accuracy improvements using uncurated data. However, these models cannot infer functional annotations for full-length protein sequences, since they are trained on pre-segmented domains and can only predict a single label. The full-sequence task is of primary importance to biological practitioners.

@@

We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score. %
%startleaveout
(See \Cref{sup:fig:prc_ec_random}, \Cref{sup:fig:prc_go_random},
\Cref{sup:fig:prc_ec_clustered}, \Cref{sup:fig:prc_go_clustered}.)
%stopleaveout

@@

To visualise this space, we selected proteins with a single leaf-level EC number and used UMAP to compress their embeddings into two dimensions \citep{umap}.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{Figures/ec_embeddings.png}%
  \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
  \label{fig:umap}
\end{figure*}
%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} We took the embeddings for all non-enzymatic sequences in the test set and projected them using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, Q8RUD6 is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). Q8RUD6 can however act as a sulfertransferase \cite{pmid17408957}.

@@

To benchmark against a common signature-based methodology, we used InterProScan to assign protein family signatures to each test sequence. We chose InterProScan for its coverage of labels as well as its use of multiple profile-based annotation methods, including HMMER and PROSITE, mentioned above. We note that while InterProScan predicts GO labels directly, it does not do so for EC labels, which is why we did not use InterProScan to benchmark our work on predicting EC labels.

We found that InterProScan was largely fairly precise, but did not recall many labels, giving it a precision of 0.937 and recall of 0.543 for an F1 score of 0.688. We note that both ProteInfer's recall at a precision of .937 is substantially higher (0.835) than InterProScan at assigning GO labels.

There are multiple caveats to these comparisons, both in terms of the trustworthiness of the annotations and how to implement the baselines. One challenge is that Swiss-Prot's annotation of GO terms is often incomplete \cite{gray2016virusmapper}. As an extreme example, \textit{Pan paniscus} (Pygmy Chimpanzee) and \textit{Pan troglodytes} (Chimpanzee) have an identical Apolipoprotein A-II protein,\footnote{Accessions P0DM95 and Q8MIQ5.}, where the first protein has 24 GO annotations, while the latter has 143 GO annotations.\footnote{This count is done using not only the set of all labels that appear in Swiss-Prot, but also any parents of those labels.} One way this is reflected in the performance of the models is that some BLAST matches that have extremely large bit-scores are not annotated identically, and thus hurt the precision of the BLAST model. Another caveat of our evaluation is that InterProScan is not capable of predicting all labels that are in our test set, lowering its recall. We ran an experiment limiting the evaluation labels that InterProScan can produce, and while recall was improved, the qualitative outcome of a relatively low recall did not change.

@@

Using Swiss-Prot to benchmark our tool against traditional alignment-based methods has distinct advantages and disadvantages. It is desirable because the data has been carefully curated by experts, and thus it contains minimal false-positives. On the other hand, many entries come from experts applying existing computational methods, including BLAST and HMM-based approaches, to identify protein function. Therefore, the data may be enriched for sequences with functions that are easily ascribable using these techniques, and thus we have limited ability to estimate the added value of using our alternative alignment-free tool. Our clustering-based train-test split helps mitigate the issue by reducing the sequence similarity between training and evaluation sequences, but it is fundamentally limited by the diversity of the underlying dataset.

@@

Our code, data, and notebooks reproducing the analyses shown in this work are available online at \url{https://github.com/google-research/proteinfer} and \url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}.
%startleaveout
\begin{acknowledgements}
%stopleaveout
%startleaveout
\end{acknowledgements}
%stopleaveout
%startleaveout
\section*{Bibliography}
\bibliography{bibliography}
\onecolumn
\newpage
%stopleaveout
%startleaveout
\subsubsection*{Label inheritance}
Some annotations, such as GO terms and EC numbers, come as directed acyclic graphics (which take the form simple trees for EC numbers). Typically in such cases the annotation provided on UniProt is the most-specific that is known. For example if a protein is known to exhibit \textit{sequence specific DNA binding} (GO:0043565). It will not separately be annotated with the ancestral term \textit{DNA binding} (GO:0003677), this is simply assumed from the ontology. Using such an annotations directly, however, is likely to be problematic in a deep learning setting. Failing to annotate an example with the parental term demands that the model predict that the example is negative for this term, which is not the effect we want.
To address this our datasets include labels for all ancestors of applied labels for EC, GO and InterPro datasets. In the case of GO we restrict these to \textit{is\_a} relationships.
\subsubsection*{Class activation mapping}
Many proteins have multiple functional properties. For example we analyse the case of the bifunctional dhfr/ts of \textit{T. gondii}. Such bifunctional enzymes are often not unique -- it is functionally advantageous for these enzymes to be fused, which facilitates channeling of substrate between their active sites. Since there are a number of such examples within Swiss-Prot, the mere existence of a TS domain in a protein is (mild) evidence for possible DHFR function. To increase the interpretability of the network we introduce a technique that is able to simplify such cases by subtracting the activations for other predicted classes during class activation mapping. \todo{cut this section right down to basics}
\subsection*{Model architecture}
To create an architecture capable of receiving a wide range of input sequences, with computational requirements determined for each inference by the length of the individual input sequence, we employed a dilated convolutional approach \citep{dilatedconvolutions}. Computation for both training and prediction in such a model can be parallelized across the length of the sequence. By training on full length proteins, in a multi-label training setting, we aimed to build networks that could  extract functional information from raw amino acid sequences. One helpful feature of this architecture is its flexibility with regards to sequence length. Natural protein sequences can vary in length by at least three orders of magnitude, but some architectures have computational requirements that scale with the maximum sequence they are capable of receiving as input, rather than the sequence being currently examined. These fixed-length approaches reduce efficiency as well as place a hard limit on the length of sequences that can be examined. 
% Extra text

%stopleaveout
%%%BOTTOMSTUFF%%%
%startleaveout
\section*{Supplement}
\beginsupplement

@@

  \includegraphics[width=\textwidth]{Figures/prc_ec_random.png}
    \includegraphics[width=\textwidth]{Figures/prc_ec_random_unzoomed.png}
  \caption{Precision/recall tradeoff for the ProteInfer CNN, an ensemble of ProteInfer CNNs, and a BLASTp search on predicting enzymatic function on a random split of Swiss-Prot.}
  \label{sup:fig:prc_ec_random}
  \includegraphics[width=\textwidth]{Figures/prc_go_random.png}
  \includegraphics[width=\textwidth]{Figures/prc_go_random_unzoomed.png}
  \caption{Precision/recall tradeoff for the ProteInfer CNN, an ensemble of ProteInfer CNNs, and a BLASTp search on predicting gene ontology labels on a random split of Swiss-Prot.}
  \label{sup:fig:prc_go_random}
  \includegraphics[width=\textwidth]{Figures/prc_ec_clustered.png}    \includegraphics[width=\textwidth]{Figures/prc_ec_clustered_unzoomed.png}
  \caption{Precision/recall tradeoff for the ProteInfer CNN, an ensemble of ProteInfer CNNs, and a BLASTp search on predicting enzymatic function on a split of Swiss-Prot ensuring large distance between test and train sequences.}
  \label{sup:fig:prc_ec_clustered}
  \includegraphics[width=\textwidth]{Figures/prc_go_clustered.png}    \includegraphics[width=\textwidth]{Figures/prc_go_clustered_unzoomed.png}
  \caption{Precision/recall tradeoff for the ProteInfer CNN, an ensemble of ProteInfer CNNs, and a BLASTp search on predicting gene ontology labels on a split of Swiss-Prot ensuring large distance between test and train sequences.}
  \label{sup:fig:prc_go_clustered}
\subsection{Stratified performance}
  \includegraphics[width=\textwidth]{Figures/ec_f1_by_training_size.png}
  \caption{Performance of EC model stratified by number of training examples available for each test example.}
  \label{sup:fig:ec_f1_by_training_size}
\subsection{Hyperparameters}
We tuned over batch size, dilation rate, filters, first dilated layer, kernel size, learning rate, number of layers, mean vs max pooling, and Adam $\beta_1$, $\beta_2$ and $\epsilon$~\citep{kingma2014adam} over a number of studies to determine the set of parameters that optimized $\texttt{F}_{max}$. We found, as in~\cite{bileschi}, that the network was relatively unresponsive to slight changes in hyperparameters, and found that many of the hyperparameters that performed well in~\cite{bileschi} also performed well for this task. We chose to keep identical hyperparameters for the EC and GO tasks across both the random and clustered splits for simplicity, and we note that parameters with good performance on the random task performed respectively well on the clustered split. 
\def\arraystretch{1.1}
\begin{tabular}{|c|c|c|c|c|}
                           & \\
                           & CNN \\
\Xhline{1pt}
concurrent batches (data parallelism) & 8 \\
batch size                 & \shortstack{40 (per each GPU) \\ Dynamic based on sequence length} \\
dilation rate              & 3 \\
filters                    & 1100  \\
first dilated layer        & 2 \\
% zero-indexed so added one to the values shown in the hparams_registry for clarity.
gradient clip              & 1 \\
kernel size                & 9 \\
learning rate              & 1.5E-3 \\
learning rate decay rate   & 0.997 \\
learning rate decay steps  & 1000 \\
learning rate warmup steps & 3000 \\
Adam $\beta_1$              & .9 \\
Adam $\beta_2$              & .999 \\
Adam $\epsilon$             & 1E-8 \\
number of ResNet layers     & 5 \\
pooling                     & mean \\
ResNet bottleneck factor    & 0.5 \\
train steps                 & 500000 \\
\caption{Hyperparameters used in convolutional neural networks. We note that hyperparameters for single-GPU training are available in \url{github.com/google-research/proteinfer/blob/master/hparams_sets.py}.}
\label{tab:nn_hparams}
\subsection{Predicting coarse-grained functional localization with CAM}
The goal of this experimental methodology is to measure whether or not we correctly order the localization of function in bifunctional enzymes. As such, first we have to identify a set of candidates for experimentation.

>>>>

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for predicting functional properties of proteins}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy Colwell}

\affil[1]{Francis Crick Institute}
\affil[2]{Google AI}
%General points -editorialise in methods. Deep learning. Motivation. Talk about curation in abstract. Metagenomes. Discussion: mention value of dataset, and the need for future work. More discussion of the existence of prediction methods from 3D. Mention that are dataset is less of a toy than most previous.
\maketitle
%TC:break Abstract
%the command above serves to have a word count for the abstract

% Subtitle from html version
% We describe an approach for predicting the functional properties of a protein from its amino acid sequences using neural networks. You can try an implementation of our technique below, which runs locally on your own device using TensorFlowJS. Read on for an explanation of how it works, and interactive figures and notebooks analysing the network.

\begin{abstract}
Predicting the function(s) of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. In our work, we instead employ deep convolutional neural networks to directly predict a variety of protein functions - EC numbers and GO terms - directly from an unaligned amino acid sequence. This performs comparably to alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces. In response, we have provided an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}.
\end {abstract}

%TC:break main
%the command above serves to have a word count for the abstract
\begin{keywords}
neural networks | protein | prediction | classification | embedding
\end{keywords}

\begin{corrauthor}
theo \at \;theo.io
\end{corrauthor}

\begin{figure}[h!]
  \centering
  \includegraphics[width=.4\columnwidth]{Figures/qr.pdf}

 \small QR code for the interactive version of this preprint at  \url{https://google-research.github.io/proteinfer/}
\end{figure}

\section*{Introduction}
Every day, more than a hundred thousand protein sequences are added to global sequence databases \cite{tremblstats}. However, these entries are of limited use to practitioners unless they are accompanied by functional annotations. While curators diligently extract annotations from the literature, assessing more than 60,000 papers each year \cite{uniprotlatest}, the time-consuming nature of this task means that only 0.03\% of publicly available protein sequences are manually annotated.
%%%FIGURETREMBL%%%
The community has a long history of using computational tools to infer protein function directly from amino acid sequence. Starting in the 1980s, methods such as BLAST \cite{blast} relied on pairwise sequence comparisons, where a query protein is assumed to have the same function as highly similar sequences that have already been annotated. Signature-based approaches were later introduced, with the PROSITE database \cite{bairoch1991prosite} cataloguing short amino acid ``motifs'' found in proteins that share a particular function. Subsequently, a crucial refinement of signature-based approaches was the development of profile hidden Markov models (HMMs) \cite{krogh1994hidden, eddyhmmreview}. These models collapse an alignment of related protein sequences into a model that provides likelihood scores for new sequences, that describe how well they fit the aligned set. Critically, profile HMMs allow for longer signatures and fuzzier matching and are currently used to update popular databases such as Interpro and Pfam \cite{pfamlatest, blum2021interpro}. Subsequent refinements have made these techniques more sensitive and computationally efficient \cite{altschul1997gapped, fasterhmmer, pfamlatest, soding2005protein, steinegger2019hh}, while their availability as web tools allows practitioners to easily incorporate them into workflows \cite{johnson2008ncbi, soding2005hhpred, potter2018hmmer, bernhofer2021predictprotein}.
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{Figures/new_the_task.pdf}%
  \caption{\textbf{Three approaches for mapping from an amino acid sequence to inferred function:}   (1) finding similar sequences in a large database of sequences with known annotation (e.g., BLAST), (2) scoring against a large database of statistical models for each family of sequences with known function (e.g. InterProScan), and (3) applying a single deep neural network trained to predict multiple output categories (e.g., this work).\label{fig:task}}
\end{figure}
These computational modelling approaches have had great impact; however, one third of bacterial proteins still cannot be annotated (even computationally) with a function \cite{phenotypes}. It is therefore worthwhile to examine how new approaches might complement existing techniques. First, current approaches conduct entirely separate comparisons for each comparator sequence or model, and thus may not fully exploit the features shared across different functional classes. An ideal classification system, for example, might have a modular ATP-binding region detector used in detection of both kinases and ABC transporters \citep{pmid12468712}. Separately modelling these targets in each family, like standard HMM approaches, increases the computational cost and may also be less accurate. In addition, the process of creating many of these signatures is not fully automated and requires considerable curatorial efforts \citep{pfamprofilebuilding, pfamrepeats}, which at present are spread across an array of disparate but overlapping signature databases \citep{blum2021interpro}.

@@

Of particular relevance to the present work is Bileschi et al. (2019) \citep{bileschi}, where it is shown that models with residual layers \citep{resnet} of dilated convolutions \citep{dilatedconvolutions} can precisely and efficiently categorise protein domains.  Dohan (2021) \cite{dohan2021kdd} provides additional accuracy improvements using uncurated data. However, these models cannot infer functional annotations for full-length protein sequences, since they are trained on pre-segmented domains and can only predict a single label. The full-sequence task is of primary importance to biological practitioners.

@@

We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score. %
%startleaveout
(See \Cref{sup:fig:with_ensemble}, \Cref{sup:fig:without_ensemble}.)
%stopleaveout

@@

To visualise this space, we selected proteins with a single leaf-level EC number and used UMAP to compress their embeddings into two dimensions \citep{umap}.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{Figures/umap2.pdf}%
  \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
  \label{fig:umap}
\end{figure*}
%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} We took the embeddings for all non-enzymatic sequences in the test set and projected them using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, Q8RUD6 is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). Q8RUD6 can however act as a sulfertransferase \cite{pmid17408957}.

@@

To benchmark against a common signature-based methodology, we used InterProScan to assign protein family signatures to each test sequence. We chose InterProScan for its coverage of labels as well as its use of multiple profile-based annotation methods, including HMMER and PROSITE, mentioned above. We note that while InterProScan predicts GO labels directly, it does not do so for EC labels, which is why we did not use InterProScan to benchmark our work on predicting EC labels. We found that InterProScan gave good precision, but within this UniProt data had lower recall, giving it a precision of 0.937 and recall of 0.543 for an F1 score of 0.688. ProteInfer's recall at a precision of .937 is substantially higher (0.835) than InterProScan at assigning GO labels.

There are multiple caveats to these comparisons, both in terms of the trustworthiness of the annotations and how to implement the baselines. One challenge is that the completeness of Swiss-Prot's GO term annotations varies \cite{jiang2014impact}. As an extreme example, \textit{Pan paniscus} (Pygmy Chimpanzee) and \textit{Pan troglodytes} (Chimpanzee) have an identical Apolipoprotein A-II protein,\footnote{Accessions P0DM95 and Q8MIQ5.}, where the first protein has 24 GO annotations, while the latter has 143 GO annotations.\footnote{This count is done using not only the set of all labels that appear in Swiss-Prot, but also any parents of those labels.} One way this is reflected in the performance of the models is that some BLAST matches that have extremely large bit-scores are not annotated identically, and thus hurt the precision of the BLAST model. It is also important to note that our model has the advantage of being specifically trained on the UniProt labelling schema upon which it is being evaluated. InterPro works quite differently, with GO terms being assigned to families, and so inconsistencies in terms of how these are assigned can explain reduced performance -- for instance InterPro families simply do not feature all of the GO terms found in UniProt. Thus these results should be seen as specific to the task of reproducing the curated results in UniProt rather than a broader objective ground truth.

@@

Using Swiss-Prot to benchmark our tool against traditional alignment-based methods has distinct advantages and disadvantages. It is desirable because the data has been carefully curated by experts, and thus it contains minimal false-positives. On the other hand, many entries come from experts applying existing computational methods, including BLAST and HMM-based approaches, to identify protein function. Therefore, the data may be enriched for sequences with functions that are easily ascribable using these techniqueswhich could limit the ability to estimate the added value of using an alternative alignment-free tool. An idealised dataset would involved training only on those sequences that have themselves been experimentally characterized, but at present  too little data exists than would be needed for a fully supervised deep-learning approach. Semi-supervised approaches that combine a smaller number of high quality experimental labels with the vast set of amino acid sequences in TrEMBL may be a productive way forward.

@@

Our code, data, and notebooks reproducing the analyses shown in this work are available online at \url{https://github.com/google-research/proteinfer} and \url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}.
%startleaveout
\begin{acknowledgements}
%stopleaveout
%startleaveout
\end{acknowledgements}
%stopleaveout
%startleaveout
\section*{Bibliography}
\bibliography{bibliography}
\onecolumn
\newpage
%stopleaveout
%startleaveout

% Extra text

%stopleaveout
%%%BOTTOMSTUFF%%%
%startleaveout
\section*{Supplement}
\beginsupplement

\subsection{implementation details}

\subsubsection*{Label inheritance}
Some annotations, such as GO terms and EC numbers, come as directed acyclic graphics (which take the form simple trees for EC numbers). Typically in such cases the annotation provided on UniProt is the most-specific that is known. For example if a protein is known to exhibit \textit{sequence specific DNA binding} (GO:0043565). It will not separately be annotated with the ancestral term \textit{DNA binding} (GO:0003677), this is simply assumed from the ontology. Using such an annotations directly, however, is likely to be problematic in a deep learning setting. Failing to annotate an example with the parental term demands that the model predict that the example is negative for this term, which is not the effect we want.
To address this our datasets include labels for all ancestors of applied labels for EC, GO and InterPro datasets. In the case of GO we restrict these to \textit{is\_a} relationships.

\subsubsection*{Class activation mapping}
Many proteins have multiple functional properties. For example we analyse the case of the bifunctional dhfr/ts of \textit{T. gondii}. Such bifunctional enzymes are often not unique -- it is functionally advantageous for these enzymes to be fused, which facilitates channeling of substrate between their active sites. Since there are a number of such examples within Swiss-Prot, the mere existence of a TS domain in a protein is (mild) evidence for possible DHFR function. To increase the interpretability of the network we introduce a technique that simplifies such cases by subtracting the activations for other predicted classes during class activation mapping.

\subsection*{Model architecture}
To create an architecture capable of receiving a wide range of input sequences, with computational requirements determined for each inference by the length of the individual input sequence, we employed a dilated convolutional approach \citep{dilatedconvolutions}. Computation for both training and prediction in such a model can be parallelized across the length of the sequence. By training on full length proteins, in a multi-label training setting, we aimed to build networks that could  extract functional information from raw amino acid sequences. One helpful feature of this architecture is its flexibility with regards to sequence length. Natural protein sequences can vary in length by at least three orders of magnitude, but some architectures have computational requirements that scale with the maximum sequence they are capable of receiving as input, rather than the sequence being currently examined. These fixed-length approaches reduce efficiency as well as place a hard limit on the length of sequences that can be examined.

@@

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated.pdf}
    \caption{EC number prediction: random split} 
    \label{with_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated_go.pdf}
    \caption{GO term prediction: random split} 
    \label{with_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated.pdf}
    \caption{EC number prediction: clustered split} 
    \label{with_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated_go.pdf}
    \caption{GO term prediction: clustered split} 
    \label{with_ensemble:d} 
  \end{subfigure} 
  \caption{Bootstrapped precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs, and ensembled ProteInfer CNNs scaled by BLAST score. }
  \label{with_ensemble} 
\end{figure}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_random_unzoomed.png}
    \caption{EC number prediction: random split} 
    \label{without_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_random_unzoomed.png}
    \caption{GO term prediction: random split} 
    \label{without_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_clustered_unzoomed.png}
    \caption{EC number prediction: clustered split} 
    \label{without_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_clustered_unzoomed.png}
    \caption{GO term prediction: clustered split} 
    \label{without_ensemble:d} 
  \end{subfigure} 
  \caption{Full precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs }
  \label{without_ensemble} 
\end{figure}
